Characterizing narrative entities with embeddings : a case study of character relationships in Les Misérables

### 2022-10-12 article/paper1.tex
write paper

In the field of \emph{Digital Humanities}, \emph{Distant Reading} tools \cite{Moretti2014} allow researchers to quickly gain knowledge on textual corpora without actually reading them. Purposes of these methods are various, but can be mainly categorized into two groups: in the first case, these methods are used in order to tag, classify, or summarize large quantities of texts, in order to quickly structure information or to deliver a speech over the whole studied corpus \cite{Underwood2017}. Methods in this case rely heavily on Big Data and make an extensive use of Machine Learning, often with the help of supervised methods. In the second case, researchers use computational methods to underline hidden structures in a small corpus or even a lone text, which helps them to refine their understanding of this corpora and reinforce stated hypotheses \cite{Eve2017}. Methods in this setting can also rely on Machine Learning, but must typically be build with more caution and attention to details: corpora are small, analyses are closer to the work, and methods must be more transparent in order to appropriately interpret results. The use of exploratory tools and unsupervised methods is also preferred in this context, as it is less desirable to base methods on information coming from large external corpora. The proposed method typically belong to this second group of methods, as it is unsupervised and work on a single document.

When a lone (or a few) textual narrative is analyzed, a particular effort is made into studying \emph{narrative entities} (characters, event, location, etc.) used by the author in her/his book \cite{Schmid2010}. Researchers are often interested by characterizing them and see how they are articulated to one another in the story. Various computational tools can help them in this tasks, to name a few, Named Entity Recognition tools \cite{Agarwal2013, Chaturvedi2017, Li2022}, Automatic Character Networks Extraction \cite{Labatut2019}, Sentiment Analysis and Topic Modeling \cite{Min2019}, Textometry \cite{Novakova2019}, and Word Embeddings \cite{Grayson2016, Heuser2017, Kerr2017}. All these methods have been used in order to explicitly show hidden structures constructed by the author in her/his work. It permits to find regularities and patterns, and can help to categorize particular narrative constructions, writing styles, or genres. These kind of methods can be great complement to classical analyses of literacy works as they allow to efficiently summarized information which is otherwise quite diffuse.

In this article, we propose a general framework in order to automatically characterize various narrative entities in a literacy work. The entire framework is exposed starting from a very wide perspective, which is how to organize the textual data, and is narrowed down to a very specific use, the study of character relationships in \emph{Les Misérables}, by Victor Hugo. Along this presentation, various choices are made to highlight a particular usage of this framework, but theses choices should be viewed as suggestion rather than rules: the real strength of this framework is its flexibility and the direction taken in this article is oriented for a particular task. To be more specific, we are interested here in using \emph{embeddings} in order to locate \emph{character relationships} alongside the vocabulary. Similarity measures can then be constructed between these words and the relationships, which can help a practitioner to characterize them. Four variations of this method are proposed, and are tested on \emph{Les Misérables}.

The idea behind this methodology comes from the field of automatic extraction and analysis of \emph{character networks} from literacy works (see \cite{Labatut2019} for a survey). When building character networks from a textual narrative, one of the most widespread methods consists in dividing the studied work into $n$ \emph{narrative units} or \emph{contexts} $u_1, \ldots, u_n$, which can be, e.g., sentences, paragraphs, or chapters, and then counting the number of units where characters co-occurred \cite{Elsner2012, Lee2012, Rochat2014, Grener2017, Min2019}. Usually, the text constituting these units is discarded and the resulting network displays edges which roughly represent an aggregated number of interactions between characters. However, by doing so, the aggregation occurs on various type of interactions and will give little information about the type of relationship which exist between characters. Various improvements where proposed in order to weight \cite{Sack2014} or sign \cite{Krishnan2015} (or both \cite{Min2019}) the edges in the character networks. A particular inspiration for the current work is the article by Min and Park \cite{Min2019}, where authors also analyzed characters in \emph{Les Misérables} by building various signed and weighted networks, with the help of Sentiment Analysis and Topic Modeling. The current framework was build by expanding this idea of refining character relationships while keeping exploring directions as wide as possible. Embeddings \cite{Incitti2023} appeared to us to be the proper tool for achieving this. As a matter of fact, with embeddings, the textual contents of units are transformed into workable mathematical objects (the vectors), usable for various tasks, while conserving a maximum of information. The framework has been further generalized in order to be applicable on various type of narrative entities, but the presented case remains the study of character relationships in \emph{Les Misérables}.

The current article is structured as followed. Section \ref{methodology} define the methodology, with section \ref{data} defining the data organization, section  \ref{unit_embeddings} describing how to embed textual units, and section \ref{entity_embeddings} deriving entity vectors lying in the same space as units. In section \ref{case_studie}, we present the methods and results for the case study of character relationship embeddings in \emph{Les Misérables}, and section \ref{conclusion} draw conclusions and perspectives about this work. All (Python) scripts and datasets used in this article, as well as extended results, can be found in the dedicated GitHub repository.\footnote{\url{https://github.com/gguex/char2char_vectors}.}

\section{Methodology}
\label{methodology}

@@

In this article, the textual narrative is divided in $n$ \emph{textual units} $u_1, \ldots, u_n$, and is represented through two tables. The first one is well known in the field of textual analysis, and consists in the $(n \times v)$ \emph{unit-word contingency table} $\mathbf{N}$, as represented by Table 1, where $v$ is the vocabulary size. In this table, each row represents an unit, each column a word, and cells $n_{ij}$ counts the number of times the word $i$ appears in the unit. Using this table typically denotes a \emph{Bag-of-Words} approach in our analyses.

\begin{table}[h]
	\scriptsize
	\begin{tabular}{|c||c|c|c|c|c|c|}
		\hline
		& aller & allumer & apercevoir & bas & bon & ... \\
		\hline
		\hline 
		$u_{101}$ & 23 & 2 & 6 & 11 & 6 & ... \\
		\hline
		$u_{102}$ & 12 & 1 & 0 & 3 & 9 & ... \\
		\hline
		$u_{103}$ & 10 & 0 & 5 & 1 & 5 & ... \\
		\hline
		$u_{104}$ & 0 & 0 & 1 & 0 & 0 & ... \\
		\hline
	\end{tabular}
	\label{cont_table}
	\caption{A snippet of the contingency table $\mathbf{N}$ extracted from \emph{les Misérables}. Rows are chapters, columns are words in the vocabulary, and cell $n_{ij}$ counts the number of time word $j$ appear in chapter $i$.}
\end{table}

The second table is the \emph{unit-entity table}, noted $\mathbf{E}$. It has a size of $(n \times p)$ where $p$ is the number of \emph{narrative entities} found in the text and cells $e_{ij}$ indicates the presence, or the count for a weighted version, of entity $j$ in unit $i$. An narrative entity, in the context of this article, can be loosely defined in order to be flexible for various types of works or analyses. It can roughly be seen as a recurring narrative object, which can contain one or more characters. For example, it can be a location, a character, a pair of characters (or even a triplet, a quadruplet, etc.), an oriented character interaction (e.g. a dialog), or even a particular recurring event containing multiple characters (e.g. a meeting). In this article, we mostly consider character and pair of characters as entities, as shown in Table 2. Note that we consider that a character or a pair of characters are present in the unit if character names (or aliases) are detected above a fixed threshold. A weighted version of this table, where $e_{ij}$ contain the number of occurrences of the entity $j$ in the unit $i$, is also possible and formulas are written accordingly.

\begin{table}[h]
	\scriptsize
	\begin{tabular}{|c||c|c|c|c|}
		\hline
		& Cosette & Thénardier & Valjean & ... \\
		\hline
		\hline 
		$u_{101}$ & 1 & 1 & 0 & ... \\
		\hline
		$u_{102}$ & 1 & 1 & 0 & ... \\
		\hline
		$u_{103}$ & 1 & 1 & 0 & ... \\
		\hline
		$u_{104}$ & 1 & 0 & 1 & ... \\
		\hline
	\end{tabular}
	\begin{tabular}{|c||c|c|c|c|}
		\hline
		& ... & Cosette-Thénardier & Cosette-Valjean & ... \\
		\hline
		\hline 
		$u_{101}$  	& ... & 1 & 0 &  ... \\
		\hline
		$u_{102}$ 	& ... & 1 & 0 & ... \\
		\hline
		$u_{103}$  	& ... & 1 & 0 & ... \\
		\hline
		$u_{104}$ 	& ... & 0 & 1 & ... \\
		\hline
	\end{tabular}
	\label{occ_table}
	\caption{A snippet of the entities table $\mathbf{E}$ extracted from \emph{les Misérables}. Rows are chapters, columns are characters (top) and character-pairs (bottom), and cell $e_{ij}$ denotes if $j$ appear in chapter $i$.}
\end{table}

This data organization already gives an orientation to the subsequent analyses and should be kept in mind of the practitioner. Textual unit are now considered as \emph{individuals} (in the statistical terminology), defined by their \emph{variables} contained in the different columns of both tables. Moreover, subsequent analyses are oriented in searching how the entity table $\mathbf{E}$ has an influence over the contingency table $\mathbf{N}$, i.e. searching which words are over-represented or under-represented knowing entities in the unit. While authors use characters in order to build the narrative, we, to a certain extent, work backward: we are searching how character appearances and interactions in the textual unit act on her/his choice of words. If the extraction method permits it, a practitioner should include all entities which she/he desire to study. Here, for example, the choice to include character-pairs along with characters is motivated by the fact that we are interested in studying character relationships. A character-pair can roughly be seen as an interaction between two characters, and this interaction should be considered as an object of its own: the presence of this interaction in a unit do not result in having a mixture of words used for each character, but rather gives a specific flavor to the unit.

@@

The use of a contingency table $\mathbf{N}$ to represent the textual resource present in the units denotes a \emph{Bag-of-Words} approach. Using this approach loses the information relative to the order of words in the units, but permits to transform a chain of characters, improper to statistical analyses, into a contingency table, a well studied mathematical object which allows the use of various kind of statistical methods. The next section shows a particular direction on how to use this table, with the help of \emph{embeddings}.

@@

Various methods can be performed on the contingency table $\mathbf{N}$ in order to extract information from it. Here, we make to choice to extract a lower dimensional, numeric representation of each units, in other words, an \emph{embedding of textual units}. \\
In section \ref{entity_embeddings}, these vectors of textual units are used as anchor points in order to also embed entities into the same space. Therefore, it is crucial that an interpretation about the directions or the regions of this embedding is possible, in order to properly interpret the localization of entity vectors (the relative position of entity vectors among themselves is generally insufficient). For that reason, we focus on embeddings of textual units which also contain \emph{vectors of words}: by examining the positions of entities relatively to word vectors, entities can be characterized. We propose two embeddings verifying this condition: Section \ref{ca_method} describes \emph{Correspondence Analysis (\textbf{CA})} and section \ref{wv_method} focuses on \emph{Pre-trained Word Vectors (\textbf{WV})}.

@@

The association between a particular unit $i$ and a word $j$ is expressed through the scalar product between their vectors $s_{ij} := \mathbf{x}^\top_i \mathbf{w}_j$, which act as a similarity measure. A positive (resp. negative) similarity denotes a over-representation (resp. under-representation) of the word $j$ in $i$, which permit to find lists of words characterizing the different units. Observe that in this article, this similarity is rather computed between a word vector and a entity vector, since the latter, as we will see in section \ref{entity_embeddings}, lies in the same space as unit vectors.

Note that vectors $\mathbf{x}_1, \ldots, \mathbf{x}_n$ obtained from \textbf{CA} reflect textual unit profile (in terms of words) regarding the mean profile (the origin in the factorial map). This analysis is thus \emph{contrastive}: it highlights unit variations among the studied text. It means that the particular tone of the whole studied text might be hidden in this analysis and only the variation around this tone will be revealed. It might lead to the situation where the (absolute) feeling experienced by the reader will not appear in this analysis, e.g. a sad character in a sad book will appear joyful if he is less sad than the mean tone. This could becomes problematic when this method is used sequentially to study multiple works: particularities of each book will be hidden. Another limitation with this approach is that the words helping the interpretation of units (and entities) are contained in the studied text. Approaches requiring to study the position of units and entities relatively to a predefined list of words (e.g., friends, enemies, family) might therefore be impossible if these words do not appear in the text.

@@

\emph{Pre-trained Word Vectors}$  $ (\textbf{WV}), based on methods such as Word2Vec \cite{Mikolov2013}, GloVe \cite{Pennington2014}, fastText \cite{Bojanowski2017}, or Bert \cite{Devlin2018} have received great attention from various fields in the last decade. They are generally obtained through a training on a very large corpora, such as Wikipedia or Common Crawl, and the resulting embedding contains a large quantity of word vectors. As shown by multiple studies (see \cite{Li2017} for a survey), these vectors are placed in order to reflect semantic and syntactic similarities between words, and are used in various applications. We will focus here on \emph{static} word embeddings, where word vectors are fixed and do not depends on their context, like Word2Vec or fastText, because we need to have a interpretable regions of the embedding space.

There exists multiple methods in order to find vectors for a \emph{group of words}, such as sentences \cite{Grener2017, Arora2017}, paragraphs \cite{Le2014}, or documents \cite{Kusner2015}, based on pre-trained word vectors. These derived vectors are often used to apply a classification or clustering algorithm on the newly embedded objects, or to query information \cite{Kusner2015, Arora2017}. In order to derive vectors for groups of words, the majority of methods use the frequencies of words in the objects, i.e. a table similar to $\mathbf{N}$, but apply various weighting scheme and normalization in order to reduce the effects of frequent words and to standardize vectors. In the present article, we use a methodology proposed in \cite{Arora2017} as it is compatible with multiple unit sizes and gives state-of-the-art results in many tasks.

Textual units vectors $\mathbf{x}_1, \ldots, \mathbf{x}_n$ are obtained through the table $\mathbf{N}$ and with the method detailed in Appendix \ref{wv_details}. The main way to interpret units with this method is to use the cosine similarity between its corresponding vector $\mathbf{x}_i$ and every word-vectors $\mathbf{w}_j$, as defined by $s_{ij} := \frac{\mathbf{x}_i^\top \mathbf{w}_j}{\sqrt{\mathbf{x}_i^\top \mathbf{x}_i \mathbf{w}_j^\top \mathbf{w}_j}}$. The cosine similarity also permits to compare units between themselves.

With the pre-trained word vector method, the unit vectors $\mathbf{x}_1, \ldots, \mathbf{x}_n$ (and entity vectors in section \ref{entity_embeddings}) lie in an \emph{absolute space} defined by the pre-trained word vectors. Comparison between different texts are therefore more pertinent in this space and comparison with words absent from the corpus can be made. However, it is still possible that all units from a given text will be located in the same region of the space, if the style of the text is particular. In this case, the list of most associated word-vectors might be similar for every unit, and the analysis will not give satisfying results. This effect is fortunately limited by the centration of unit-vectors which occurs in the method describe in Appendix \ref{wv_details}.

@@

The main goal of this article is not to analyze units, but rather the entities, i.e. the $p$ columns of table $\mathbf{E}$. While we used the table $\mathbf{N}$ to build our embeddings of units, we now use the table $\mathbf{E}$ in order to build the entity vectors $\mathbf{y}_1, \ldots, \mathbf{y}_p$ relatively to unit vectors $\mathbf{x}_1, \ldots, \mathbf{x}_n$.

Two propositions of method are made: the \emph{centroids method} (\textbf{CENT}) is describe in section \ref{centroid} and the \emph{regressions method} (\textbf{REG}) is explained in section \ref{regression}. Both methods work with the aforementioned embeddings of units.

@@

This method is the most trivial and is based on the following intuition: an entity is characterized by all units in which it appears. In other words, we can define the vector $\mathbf{y}_k$ for character-object object $k$ as
\begin{equation}
\mathbf{y}_k = \sum_{i=1}^n \frac{e_{ik}}{e_{\bullet k}} \mathbf{x}_i
\end{equation}
i.e. the center of mass, or \emph{centroid}, of the units containing the entity, weighted by the relative weight $\frac{e_{ik}}{e_{\bullet k}}$. This way of building entity vectors is closely related to the treatment of \emph{supplementary variables} found in \textbf{CA}: these variables do not act in the choice of factorial axis, but can still be represented afterward. However, by contrast, entity vectors are not dilated after computing centroids, which means that they lie in the same space as units (row). \\
An important remark about the centroid method, is that entity vectors positions are \emph{additive}, which means that we have
\begin{equation}
e_{ik} = \sum_{g \in \mathcal{G}} e_{ig}, \forall i \implies \mathbf{y}_k = \sum_{g \in \mathcal{G}} \frac{e_{\bullet g}}{e_{\bullet k}} \mathbf{y}_g,
\end{equation}
where $\mathcal{G}$ is a subset of entities. This property can be interpreted as followed: if a character $k$ can be divided among different situations $g$ (the character alone, the character in interaction with another character, etc.), the character vector $\mathbf{y}_k$ is in fact the centroid of all vectors $\mathbf{y}_g$ of these situations. This is not necessary an undesirable property, but it implies that the specificities of the lone character might be hidden if he is often registered in an interaction. By contrast, if we consider that an interaction between two characters is an \emph{emerging situation}, unrelated to prior behaviors of characters, the regressions method described in the next section seems more appropriate.

@@

As the number of entities (i.e. predictors) might be very large, it is a good idea to add a $L^2$ regularization term in the objective function. Moreover, the quadratic error rate should also be weighted by the number of tokens in each unit. Including all this, we find the solution for our character-based vectors $\mathbf{y}_0, \mathbf{y}_1, \ldots, \mathbf{y}_p$ contained in rows of $\mathbf{Y}$ with 
\begin{equation}
\mathbf{Y} = (\widetilde{\mathbf{E}}^\top \textbf{Diag}(\mathbf{f}) \widetilde{\mathbf{E}} + \lambda \mathbf{I}_{(p+1)})^{-1} \widetilde{\mathbf{E}}^\top \textbf{Diag}(\mathbf{f}) \mathbf{X}, \label{reg_sol}
\end{equation}
where $\textbf{Diag}(\mathbf{f})$ is the diagonal matrix containing weights of units $\mathbf{f} = \left( \frac{n_{i \bullet}}{n_{\bullet \bullet}} \right)$, $\lambda > 1$ is the regularization coefficient, and $\mathbf{I}_{(p+1)}$ is the identity matrix of size $((p+1) \times (p+1))$.

>>>>

In the field of Digital Humanities, \emph{Distant Reading} tools \cite{Moretti2014} allow researchers to quickly gain knowledge on textual corpora without actually reading them. Purposes of these methods are various, but can be mainly categorized into two groups: in the first case, these methods are used in order to tag, classify, or summarize large quantities of documents, in order to quickly structure information or to deliver a speech over the whole studied corpus \cite{Underwood2017}. Methods in this case rely heavily on Big Data and make an extensive use of Machine Learning, often with the help of supervised methods. In the second case, researchers use computational methods to underline hidden structures in a small corpus or even a single document, which helps them to refine their understanding of this corpus or to validate hypotheses \cite{Eve2017}. Methods in this setting can also rely on Machine Learning, but must typically be build with more caution and attention to details: corpora are smaller, analyses are closer to the work, and methods must transparent in order to appropriately interpret results. The use of exploratory tools and unsupervised methods is also preferred in this context, as it is less desirable to base methods on information coming from large external corpora. The proposed method in this article typically belongs to the second group, as it is unsupervised and can be applied on a single document.

When a single (or a few) \emph{literary work} is analyzed, a common practice is to study \emph{narrative entities} (characters, event, location, etc.) used by the author in her/his book \cite{Schmid2010}. Researchers are frequently interested in characterizing them and seeing how they are articulated to one another in the story. Various computational tools can help them in this tasks, to name a few, Named Entity Recognition tools \cite{Agarwal2013, Chaturvedi2017, Li2022}, Automatic Character Networks Extraction \cite{Labatut2019}, Sentiment Analysis and Topic Modeling \cite{Min2019}, Textometry \cite{Novakova2019}, and Word Embeddings \cite{Grayson2016, Heuser2017, Kerr2017}. All these methods have been used in order to explicitly show hidden structures constructed by the author in her/his work. It permits to find regularities and patterns, and can help to categorize particular narrative constructions, writing styles, or genres. These kind of methods can be great complement to classical analyses of literary works as they allow to efficiently summarized information which is otherwise quite diffuse.

In this article, we propose a general framework in order to automatically characterize various narrative entities in a literary work. The entire framework is exposed starting from a wide perspective, which is how to organize the textual data, and is narrowed down to a specific use, the study of character relationships in \emph{Les Misérables}, by Victor Hugo. Along this presentation, various choices are made to highlight a particular usage of this framework, but theses choices should be viewed as suggestion rather than rules: the real strength of this framework is its flexibility and the direction taken in this article is oriented for a defined task. To be more specific, we will show how to use \emph{embeddings} in order to locate \emph{character relationships} alongside the vocabulary. An association measure can then be constructed between these words and the relationships, which can help a practitioner to characterize the latters. Four variations of this method are proposed, and are tested on \emph{Les Misérables}.

The idea behind this framework comes from the field of automatic extraction and analysis of \emph{character networks} from literacy works (see \cite{Labatut2019} for a survey). When building character networks from a textual narrative, one of the most widespread methods consists in dividing the studied work into $n$ \emph{narrative units} or \emph{contexts} $u_1, \ldots, u_n$, which can be, e.g., sentences, paragraphs, or chapters, and then counting the number of units where characters co-occurred \cite{Elsner2012, Lee2012, Rochat2014, Grener2017, Min2019}. Usually, the text constituting these units is discarded and the resulting network displays edges which roughly represent an aggregated number of interactions between characters. However, by doing so, the aggregation occurs on various type of interactions and gives little information about the type of relationship which exist between characters. Various improvements where proposed in order to weight \cite{Sack2014} or sign \cite{Krishnan2015} (or both \cite{Min2019}) the edges in the character networks. A particular inspiration for the current work is the article by Min and Park (2019) \cite{Min2019}, where authors also analyzed characters in \emph{Les Misérables} by building various signed and weighted networks, with the help of Sentiment Analysis and Topic Modeling. The current framework was build by expanding this idea of refining character relationships by formalizing the data structure and keeping directions of exploration as wide as possible. Embeddings \cite{Incitti2023} appeared to us to be the proper tool for achieving this. As a matter of fact, with embeddings, the textual contents of units are transformed into workable mathematical objects (the vectors), usable for various tasks, while conserving a maximum of information. The framework has been further generalized in order to be applicable on various type of narrative entities, but the presented case remains the study of character relationships in \emph{Les Misérables}.

The current article is structured as followed. Section \ref{methodology} define the framework, with section \ref{data} defining the data organization, section  \ref{unit_embeddings} describing how to embed textual units, and section \ref{entity_embeddings} deriving entity vectors lying in the same space as units. In section \ref{case_studie}, we present the specific methodology and results for the case study of character relationships in \emph{Les Misérables}, and section \ref{conclusion} draw conclusions and perspectives about this work. All (Python) scripts and datasets used in this article, as well as extended results, can be found in the dedicated GitHub repository.\footnote{\url{https://github.com/gguex/char2char_vectors}.}

\section{Framework}
\label{methodology}

@@

In this article, a textual narrative is divided in $n$ \emph{textual units} $u_1, \ldots, u_n$, and is represented through two tables. The first one is well known in the field of textual analysis, and consists in the $(n \times v)$ \emph{unit-word contingency table} $\mathbf{N}$, as represented by Table 1, where $v$ is the vocabulary size. In this table, each row represents an unit, each column a word, and cells $n_{ij}$ counts the number of times the word $i$ appears in the unit. Using this table typically denotes a \emph{Bag-of-Words} approach in our analyses.

\begin{table}[h]
	\scriptsize
	\begin{tabular}{|c||c|c|c|c|c|c|}
		\hline
		& aller & allumer & apercevoir & bas & bon & ... \\
		\hline
		\hline 
		$u_{101}$ & 23 & 2 & 6 & 11 & 6 & ... \\
		\hline
		$u_{102}$ & 12 & 1 & 0 & 3 & 9 & ... \\
		\hline
		$u_{103}$ & 10 & 0 & 5 & 1 & 5 & ... \\
		\hline
		$u_{104}$ & 0 & 0 & 1 & 0 & 0 & ... \\
		\hline
	\end{tabular}
	\label{cont_table}
	\caption{A snippet of the unit-word contingency table $\mathbf{N}$ extracted from \emph{les Misérables}. Rows are chapters, columns are words in the vocabulary, and cell $n_{ij}$ counts the number of time word $j$ appear in chapter $i$.}
\end{table}

The second table is the \emph{unit-entity table}, noted $\mathbf{E}$. It has a size of $(n \times p)$ where $p$ is the number of \emph{narrative entities} found in the text and cells $e_{ij}$ indicates the presence, or the count for a weighted version, of entity $j$ in unit $i$. An narrative entity, in the context of this article, can be loosely defined in order to be flexible for various types of texts or analyses. It can roughly be seen as a recurring object with some importance in the narration. For example, it can be a location, an object, a character, a pair of characters (or even a triplet, a quadruplet, etc.), an oriented character interaction (e.g. a dialog), or even a particular recurring event containing multiple characters (e.g. a meeting). In this article, we mostly consider character and pairs of characters as entities, as shown in Table 2. Note that in the present case, we consider that a character or a pair of characters are present in the unit if character names (or aliases) are detected above a fixed threshold. A weighted version of this table, where $e_{ij}$ contains the number of occurrences of the entity $j$ in the unit $i$, is also possible. However, equations presented in this article are written for the presence/absence version.

\begin{table}[h]
	\scriptsize
	\begin{tabular}{|c||c|c|c|c|}
		\hline
		& Cosette & Thénardier & Valjean & ... \\
		\hline
		\hline 
		$u_{101}$ & 1 & 1 & 0 & ... \\
		\hline
		$u_{102}$ & 1 & 1 & 0 & ... \\
		\hline
		$u_{103}$ & 1 & 1 & 0 & ... \\
		\hline
		$u_{104}$ & 1 & 0 & 1 & ... \\
		\hline
	\end{tabular}
	\begin{tabular}{|c||c|c|c|c|}
		\hline
		& ... & Cosette-Thénardier & Cosette-Valjean & ... \\
		\hline
		\hline 
		$u_{101}$  	& ... & 1 & 0 &  ... \\
		\hline
		$u_{102}$ 	& ... & 1 & 0 & ... \\
		\hline
		$u_{103}$  	& ... & 1 & 0 & ... \\
		\hline
		$u_{104}$ 	& ... & 0 & 1 & ... \\
		\hline
	\end{tabular}
	\label{occ_table}
	\caption{A snippet of the unit-entity table $\mathbf{E}$ extracted from \emph{les Misérables}. Rows are chapters, columns are characters (top) and character-pairs (bottom), and cell $e_{ij}$ denotes if $j$ appear in chapter $i$.}
\end{table}

This data organization already gives an orientation to subsequent analyses, and should be kept in mind of the practitioner. Textual unit are now considered as \emph{individuals} (in the statistical terminology), defined by their \emph{variables} contained in the different columns of both tables. Moreover, subsequent analyses are oriented in searching how the unit-entity table $\mathbf{E}$ has an influence over the unit-word table $\mathbf{N}$, i.e. searching which words are over-represented or under-represented knowing entities in the unit. While an authors uses characters in order to build her/his narrative, we, to a certain extent, work backward: we are searching how character appearances and interactions in the textual unit act on her/his choice of words. If the extraction method permits it, a practitioner should include all entities which she/he desires to study. Here, for example, the choice to include character-pairs along with characters is motivated by the fact that we are interested in studying character relationships. A character-pair can roughly be seen as an interaction between two characters, and this interaction should be considered as an object of its own: the presence of this interaction in a unit do not result in having a mixture of words used for each character, but rather gives a specific flavor to the unit.

@@

The use of a contingency table $\mathbf{N}$ to represent the textual resource present in the units denotes a \emph{Bag-of-Words} approach. Using this approach loses the information relative to the order of words in the units, but permits to transform a chain of characters, improper to statistical analyses, into a contingency table, a well studied mathematical object which allows the use of various kind of computational methods. The next section shows a particular direction on how to use this table, with the help of embeddings.

@@

Various methods can be performed on the contingency table $\mathbf{N}$ in order to extract information from it. Here, we make to choice to extract a lower dimensional, numeric representation of each unit, in other words, a \emph{textual unit vector} located in an \emph{embedding space}. \\
In section \ref{entity_embeddings}, these vectors of textual units are used as anchor points in order to also embed entities into the same space. Therefore, it is crucial that an interpretation about the directions or the regions of this embedding space is possible, in order to properly interpret the localization of entity vectors (the relative position of entity vectors among themselves is generally insufficient). For that reason, we focus on embeddings of textual units which also contain \emph{vectors of words}: by examining the positions of entities relatively to word vectors, entities can be characterized. We propose two embeddings verifying this condition: Section \ref{ca_method} describes \emph{Correspondence Analysis (\textbf{CA})} and section \ref{wv_method} focuses on \emph{Pre-trained Word Vectors (\textbf{WV})}.

@@

An \emph{association score} between a particular unit $i$ and a word $j$ is expressed through the scalar product between their vectors 
\begin{equation}
a_{ij} := \mathbf{x}^\top_i \mathbf{w}_j.
\end{equation}
A positive (resp. negative) association score denotes a over-representation (resp. under-representation) of the word $j$ in $i$, which permit to find lists of words characterizing the different units. Observe that in this article, this association score is rather computed between a word vector and a entity vector, since the latter, as we will see in section \ref{entity_embeddings}, lies in the same space as unit vectors. We could also track how units (or entities) are dissimilar to each others by using this time the Euclidean distance between vectors.

Note that vectors $\mathbf{x}_1, \ldots, \mathbf{x}_n$ obtained from \textbf{CA} reflect textual unit profile (in terms of words) regarding the mean profile (the origin in the factorial map). This analysis is thus \emph{contrastive}: it highlights unit variations among the studied text. It means that the particular tone of the whole studied text might be hidden in this analysis and only the variation around this tone will be revealed. It might lead to the situation where the (absolute) feeling experienced by the reader will not appear in this analysis, e.g., a sad character in a sad book might appear joyful if he is less sad than the mean tone. This can become problematic when this method is used sequentially to study multiple works: particularities of each book will be hidden. Another limitation with this approach is that the words helping the interpretation of units (and entities) are contained in the studied text. Approaches requiring to study the position of units and entities relatively to a predefined list of words (e.g., friends, enemies, family) might therefore be impossible if these words do not appear in the text.

@@

\emph{Pre-trained Word Vectors}$  $ (\textbf{WV}), based on methods such as Word2Vec \cite{Mikolov2013}, GloVe \cite{Pennington2014}, fastText \cite{Bojanowski2017}, or Bert \cite{Devlin2018} have received great attention from various fields in the last decade. They are generally obtained through a training on a very large corpora, such as Wikipedia or Common Crawl, and the resulting embedding contains a large quantity of word vectors. As shown by multiple studies (see \cite{Li2017} for a survey), these vectors are placed in order to reflect semantic and syntactic relationships between words, and are used in various applications. We focus here on \emph{static} word embeddings, where word vectors are fixed and do not depends on their context, obtained by, e.g., fastText. The reason is that we need to have interpretable regions in an unchanging embedding space.

Multiple methods exist which use pre-trained word vectors in order to derive vectors for a \emph{group of words}, such as sentences \cite{Grener2017, Arora2017}, paragraphs \cite{Le2014}, or documents \cite{Kusner2015}. These derived vectors are often used to apply a classification or clustering algorithm on the newly embedded objects, or to query information \cite{Kusner2015, Arora2017}. In order to derive these vectors, the majority of methods use frequencies of words found in objects, i.e. a table similar to $\mathbf{N}$, but apply various weighting scheme and normalization in order to reduce the effects of frequent words and to standardize vectors. In the present article, we use a methodology proposed in \cite{Arora2017} as it is compatible with multiple unit sizes and gives good results in many tasks. Thus, textual units vectors $\mathbf{x}_1, \ldots, \mathbf{x}_n$ are obtained through the table $\mathbf{N}$ and with the method detailed in Appendix \ref{wv_details}.

An \emph{association score} can again be computed between an unit (or an entity) vector $\mathbf{x}_i$ and word vector $\mathbf{w}_j$ through the cosine similarity, defined by 
\begin{equation}
a_{ij} := \frac{\mathbf{x}_i^\top \mathbf{w}_j}{\sqrt{\mathbf{x}_i^\top \mathbf{x}_i \mathbf{w}_j^\top \mathbf{w}_j}}
\end{equation}
Note that, with word vectors, this cosine similarity also permits to compare units (or entities) between themselves.

With the pre-trained word vector method, the unit vectors $\mathbf{x}_1, \ldots, \mathbf{x}_n$ (and entity vectors in section \ref{entity_embeddings}) lie in an \emph{absolute space} defined by the pre-trained word vectors. Comparison between different texts are therefore more pertinent and associations with words absent from the corpus can be made. However, it is possible that all units from a given text will be located in the same region of the space if the vocabulary used in it is very specific. In this case, the list of most associated word vectors might be similar for every unit, and the analysis will not give satisfying results. This effect is fortunately limited by the centration of unit vectors which occurs in the method described in Appendix \ref{wv_details}.

@@

The main goal of this article is not to analyze units, but rather entities, i.e., the $p$ columns of table $\mathbf{E}$. While we use the table $\mathbf{N}$ to build embeddings of units, we utilize the table $\mathbf{E}$ in order to build the entity vectors $\mathbf{y}_1, \ldots, \mathbf{y}_p$ relatively to unit vectors $\mathbf{x}_1, \ldots, \mathbf{x}_n$. Two propositions of methods are made: the \emph{centroids method} (\textbf{CENT}), described in section \ref{centroid}; and the \emph{regressions method} (\textbf{REG}), explained in section \ref{regression}. Both methods can be combined with the embeddings of units defined in the previous section.

@@

This method is the most trivial and is based on the following intuition: an entity is characterized equally by all units in which it appears. In other words, we can define the vector $\mathbf{y}_k$ for entity $k$ as
\begin{equation}
\mathbf{y}_k = \sum_{i=1}^n f_i e_{ik} \mathbf{x}_i
\end{equation}
where $f_i = \frac{n_{i \bullet}}{n_{\bullet \bullet}}$ is the relative weight of unit $i$. $\mathbf{y}_k$ indicates the center of mass, or \emph{centroid}, of the units containing the entity. This way of building entity vectors is closely related to the treatment of \emph{supplementary variables} found in \textbf{CA}: these variables do not act in the choice of factorial axis, but can still be represented afterward. However, by contrast, entity vectors are not dilated after computing centroids, which means that they lie in the same space as units (row). \\
An important remark about the centroid method, is that entity vectors positions are \emph{additive}, i.e. we have
\begin{equation}
e_{ik} = \sum_{g \in \mathcal{G}} e_{ig}, \forall i \implies \mathbf{y}_k = \sum_{g \in \mathcal{G}} \mathbf{y}_g,
\end{equation}
where $\mathcal{G}$ is a subset of entities. This property can be interpreted as followed: if a character $k$ can be divided among different situations $g$ (the character alone, the character in interaction with another character, etc.), the character vector $\mathbf{y}_k$ is in fact the sum of all vectors $\mathbf{y}_g$ of these situations. This is not necessary an undesirable property, but it implies that the specificities of the lone character might be hidden if he is often registered in an interaction. By contrast, if we consider that an interaction between two characters is an \emph{emerging situation}, unrelated to prior behaviors of characters, the regressions method described in the next section seems more appropriate.

@@

As the number of entities (i.e. predictors) might be very large, it is a good idea to add a $L^2$ regularization term in the objective function. Moreover, the quadratic error rate should also be weighted by the number of tokens in each unit. Including all this, we find the solution for our character-based vectors $\mathbf{y}_0, \mathbf{y}_1, \ldots, \mathbf{y}_p$ contained in rows of $\mathbf{Y}$ with 
\begin{equation}
\mathbf{Y} = (\widetilde{\mathbf{E}}^\top \textbf{Diag}(\mathbf{f}) \widetilde{\mathbf{E}} + \lambda \mathbf{I}_{(p+1)})^{-1} \widetilde{\mathbf{E}}^\top \textbf{Diag}(\mathbf{f}) \mathbf{X}, \label{reg_sol}
\end{equation}
where $\textbf{Diag}(\mathbf{f})$ is the diagonal matrix containing weights of units $\mathbf{f} = \left( f_i \right)$, $\lambda > 1$ is the regularization coefficient, and $\mathbf{I}_{(p+1)}$ is the identity matrix of size $((p+1) \times (p+1))$.
